---
title: "Simulating time-to-event trials in parallel"
author: "Cole Manschot, Nan Xiao"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating time-to-event trials in parallel}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates the ability to implement `sim_fixed_n()` using user-defined backends to parallelize simulations. We will consider the backends supported by the [future](https://future.futureverse.org/) framework.

The backends supported by the future package include:

* `sequential` - default and non-parallel backend.
* `multisession` - uses multiple background R sessions on a single machine.
* `multicore` - uses multiple forked R processes on a single non-Windows machine outside of RStudio.
* `cluster` - supports external R sessions across multiple machines.

You can also choose other backend types supported by additional future extension packages, such as the HPC job scheduler backends from future.batchtools.

The function `sim_fixed_n()` provides a simulation workflow for a two-arm trial with a single endpoint.
We can vary the parameters of the trial using different functions outlined in the documentation.
This function now provides users the opportunity to implement their simulations using the previously described parallel backends to accelerate the computation.

The function `sim_gs_n()` which simulates group sequential designs under a fixed sample size also supports the use of user-defined backends to parallelize simulations in a similar manner.

## Background

Without specifying a backend, `sim_fixed_n()` will execute sequentially.
The sequential execution will run all `n_sim` iterations within the same process or session of R.
In order to execute in parallel, we must define the environment prior to calling the function.
Setting your seed prior to calling the function will ensure that the results are reproducible.

Suppose you want to investigate the duration of a trial under two possible enrollments strategies.
Both enrollments are piecewise, but have varying durations and rates.

```{r dependencies}
library(simtrial)
library(future)
library(doFuture)
```

```{r set-plan-sequential, echo=FALSE}
plan("sequential") # Ensure that the backend is sequential
```

```{r enrollments, fig.height=4, fig.width=6, fig.align="center"}
set.seed(1)

n <- 5000
enroll_rate1 <- data.frame(rate = c(5, 20, 10), duration = c(100, 150, 150))
enroll_rate2 <- data.frame(rate = c(10, 15, 30), duration = c(150, 175, 75))
x1 <- rpwexp_enroll(n = n, enroll_rate = enroll_rate1)
x2 <- rpwexp_enroll(n = n, enroll_rate = enroll_rate2)
```

We see that *Enrollment 2* enrolls individuals more quickly than *Enrollment 1* at onset.
Later, *Enrollment 1* will outpace *Enrollment 2* before eventually being overtaken again.
As such, we want to consider how the duration of the study changes under these enrollments.

## The sequential run

Naively, we can execute these simulations sequentially.
We set the target of a total enrollment of 3000 individuals with the trial ending after observing 700 events.
We use `timing_type = 2` to return the correct trial duration.

*Note:* We manually set the number of threads to be used by {data.table} operations to 1.
This is purely for the sake of comparing the runtime to the parallel run performed later in this vignette.
If you were running simulations sequentially, you would want {data.table} to take advantage of parallel processing.

```{r confirm-sequential}
data.table::setDTthreads(threads = 1)
set.seed(1)

n_sim <- 200

start_sequential <- proc.time()

seq_result1 <- sim_fixed_n(
  n_sim = n_sim,
  sample_size = 3000,
  target_event = 700,
  enroll_rate = enroll_rate1,
  timing_type = 2 # Time until targeted event count achieved
)

seq_result2 <- sim_fixed_n(
  n_sim = n_sim,
  sample_size = 3000,
  target_event = 700,
  enroll_rate = enroll_rate2,
  timing_type = 2 # Time until targeted event count achieved
)

duration_sequential <- proc.time() - start_sequential
```

A message automatically appears in the console that indicates what backend is being used for processing.

The calls to `proc.time()` allow us to evaluate the computation time of these procedures.
This function provides three outputs, we will focus on user and elapsed time.
User time represents the CPU time spent evaluating the function and elapsed time represents the "wall clock" time spent by the end user waiting for the results.

```{r sequential-time}
print(duration_sequential)
```

We can see that the CPU time is `r sprintf("%.2f", duration_sequential[[1]])` and the elapsed time is `r sprintf("%.2f", duration_sequential[[3]])` seconds.
These provide our baseline for the computation time.

As you may have anticipated, we see that for a lower number of events, enrollment 2 has a shorter average duration of
`r sprintf("%.1f", mean(seq_result2$duration))`
over that of enrollment 1, which is
`r sprintf("%.1f", mean(seq_result1$duration))`.

## Setting up a parallel backend

If we increased the number of simulations for each enrollment, we can expect the time to run our simulations to increase.
Furthermore, as we vary and increase the number of parameter inputs that we consider, we expect the simulation process to continue to increase in duration.
To help combat the growing computational burden, we can run these simulations in parallel using the `multisession` backend available to us in `plan()`.

We can adjust the default number of cores with the function `parallelly::availableCores()`.
The multisession backend will automatically use all available cores by default, but here we will use two.
To initialize our backend, we change our plan.

```{r multisession1}
plan(multisession, workers = 2)
```

## Execution in parallel

Once we have configured the backend details, we can execute the same code as before to automatically distribute the `n_sim` simulations across the available cores.

*Note:* We do not have to worry about setting `data.table::setDTthreads(threads = 1)` for the parallel processes spawned by `sim_fixed_n()` below because {data.table} "automatically switches to single threaded mode upon fork" (from `?data.table::setDTthreads`).
[^parallel-dt-caveats]

[^parallel-dt-caveats]:
This can get complex quick.
The behavior of parallel computing is affected by your operating system (e.g. Windows) and editor (e.g. RStudio).
If you need to perform accurate benchmarking, you will need to do your own due diligence.
We recommend adding `data.table::setDTthreads(threads = 1)` to your `~/.Rprofile` to have the best chance of preventing {data.table} multithreading from affecting your benchmarking results.

```{r confirm-multisession}
set.seed(1)

start_parallel <- proc.time()

par_result1 <- sim_fixed_n(
  n_sim = n_sim,
  sample_size = 3000,
  target_event = 700,
  enroll_rate = enroll_rate1,
  timing_type = 2 # Time until targeted event count achieved
)

par_result2 <- sim_fixed_n(
  n_sim = n_sim,
  sample_size = 3000,
  target_event = 700,
  enroll_rate = enroll_rate2,
  timing_type = 2 # Time until targeted event count achieved
)

duration_parallel <- proc.time() - start_parallel
```

```{r time-parallel}
print(duration_parallel)
```

We can see that the CPU time is `r sprintf("%.2f", duration_parallel[[1]])` and the elapsed time is `r sprintf("%.2f", duration_parallel[[3]])` seconds.
The user time here appears to be drastically reduced because of how R keeps track of time; the time used by the parent process and not the children processes are reported for the user time.
Therefore, we compare the elapsed time to see the real-world impact of the parallelization.

To change the implementation back to a sequential backend, we simply use what is below.

```{r plan-sequential}
plan(sequential)
```

We can also verify that the simulation results are identical because of setting a seed and that the backend type will not affect the results. Below, it is clear that the results from our sequential and multisession backends match completely.

```{r compare-results}
all.equal(seq_result1, par_result1)
all.equal(seq_result2, par_result2)
```

*Note:* A parallel implementation may not always be faster than a serial implementation.
If there is substantial overhead associated with executing in parallel, sequential evaluation may be faster.
For a low number of simulations or available cores, it may be preferable to continue computation in serial rather than parallel.
We leave it to the end user to determine this difference based on the resources available to them.
